---
title: "PCA_usub"
author: "Summer Blanco"
date: "2026-08-17"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

# [Setup]{.underline}

```{r setup}
library(tidyverse)
setwd("/home/FM/sblanco/GWAS_8072026/usub/05_filtering")
```

# [Read in data]{.underline}

Eigenvec and eigenval files generated from a filtered VCF file using
Plink. VCF filters included:\
QUAL=30; MIN_DEPTH=10; MAX_DEPTH=250; MAF=0.10; MISS=0.75. Filtered VCF
had 10,962 SNPs (aligned to uflo ref genome). After pruning, 17,737 of
19,105 variants were removed, leaving 1,368 remaining variants.

```{r data}
pca <-read_table("usub_75_pruned.eigenvec", col_names = FALSE)
eigenval <- scan("usub_75_pruned.eigenval")
```

# [Clean data]{.underline}

## *Remove nuisance column and set names*

```{r clean}
pca <- pca[,-1]
names(pca)[1] <- "ind" 
names(pca)[2:ncol(pca)] <- paste0("PC", 1:(ncol(pca)-1))
```

## *Assign labels*

### **Species labels**

```{r names}
spp <- rep(NA, length(pca$ind))
spp[grep("UFLO", pca$ind)] <- "Usnea florida"
spp[grep("USUB", pca$ind)] <- "Usnea subfloridana"
```

### Locations labels

```{r location}
loc <- rep(NA, length(pca$ind)) 
loc[grep("11_0", pca$ind)] <- "Russia" 
loc[grep("09_", pca$ind)] <- "Latvia" 
loc[grep("UFLO_0", pca$ind)] <- "Latvia" 
loc[grep("10_0", pca$ind)] <- "Latvia" 
loc[grep("08_", pca$ind)] <- "Latvia" 
loc[grep("14", pca$ind)] <- "Belarus" 
loc[grep("FR", pca$ind)] <- "Austria" 
loc[grep("04_0", pca$ind)] <- "Estonia" 
loc[grep("05_", pca$ind)] <- "Estonia" 
loc[grep("06_", pca$ind)] <- "Estonia" 
loc[grep("06_", pca$ind)] <- "Estonia"
loc[grep("FG", pca$ind)] <- "Austria"
```

## Remake data.frame

```{r tibble}
pca <- as_tibble(data.frame(pca, spp))
```

# [Plot variance]{.underline}

## *Convert to percentage variance explained*

```{r calcvariance}
pve <- data.frame(PC = 1:20, pve = eigenval/sum(eigenval)*100)
```

## *Plot pve*

```{r plotPC}
ggplot(pve, aes(PC, pve)) + geom_bar(stat = "identity") + ylab("Percentage variance explained") + theme_light()
```

## *Calculate the cumulative sum of the percentage variance explained*

```{r cumsum}
cumsum(pve$pve)
```

> PC1 & PC2 explain \~17% of the variance observed

# [Plot PCA]{.underline}

## *By species*

```{r plotPCAspecies}
ggplot(pca, aes(PC1, PC2, shape = spp, col=spp)) + geom_point(size = 3) + coord_equal() + theme_light() + xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) + xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) + ylab(paste0("PC2 (", signif(pve$pve[2], 3), "%)"))
```

## *By location*

```{r plotPCAlocation}
ggplot(pca, aes(PC1, PC2, shape = spp, col=loc)) + geom_point(size = 3) + scale_colour_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2")) + coord_equal() + theme_light() + xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) + xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) + ylab(paste0("PC2 (", signif(pve$pve[2], 3), "%)"))
```

> Some slight structure based on region X species?
