---
title: "VCF QC — Usnea Resequencing Subset"
author: "S. Blanco"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Setup

Load tidyverse and set the working directory.

```{r load-packages}
# install.packages("tidyverse")  # uncomment if not already installed
library(tidyverse)
```

```{r set-wd}
setwd("/home/FM/sblanco/GWAS_8072026/usub/04_freebayes/QC")
```

QC visualizations for a subset of the VCF from 35 *Usnea* resequencing samples.
53,653 SNPs were randomly subsampled out of 4,496,909 SNPs for QC purposes.

## Variant Quality

Phred quality score for each site. ~24,000 SNPs have QUAL < 30 (~50%).

```{r variant-quality}
var_qual <- read_delim("usub_vcf_subset.lqual", delim = "\t",
                        col_names = c("chr", "pos", "qual"), skip = 1)

summary(var_qual$qual)

a <- ggplot(var_qual, aes(qual)) +
  geom_histogram(fill = "dodgerblue1", colour = "black", alpha = 0.3)
a + theme_light() + xlim(0, 100000)
```

> Decision: set min QUAL to 30.

## Variant Mean Depth

Mean number of reads mapped to each position.

```{r variant-depth}
var_depth <- read_delim("usub_vcf_subset.ldepth.mean", delim = "\t",
                         col_names = c("chr", "pos", "mean_depth", "var_depth"), skip = 1)

a <- ggplot(var_depth, aes(mean_depth)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3)
a + theme_light()

summary(var_depth$mean_depth)

a + theme_light() + xlim(0, 500)
```

> Decision: set min depth to 10, max depth to ~2x mean (124.52 * 2 ≈ 250).
> This handles outliers that could reflect paralogous or repetitive regions.

## Variant Missingness

Proportion of individuals lacking a genotype at each call site.

```{r variant-missingness}
var_miss <- read_delim("usub_vcf_subset.lmiss", delim = "\t",
                        col_names = c("chr", "pos", "nchr", "nfiltered", "nmiss", "fmiss"), skip = 1)

a <- ggplot(var_miss, aes(fmiss)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3)
a + theme_light()

summary(var_miss$fmiss)
```

> Typically 75–95% completeness is used as a threshold. Going with 90% (MISS = 0.10).

## Minor Allele Frequency

```{r maf}
var_freq <- read_delim("usub_vcf_subset.frq", delim = "\t",
                        col_names = c("chr", "pos", "nalleles", "nchr", "a1", "a2"), skip = 1)

var_freq$maf <- var_freq %>%
  select(a1, a2) %>%
  apply(1, function(z) min(z))

a <- ggplot(var_freq, aes(maf)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3)
a + theme_light()

summary(var_freq$maf)
```

> A large number of variants have low-frequency alleles. Will set MAF threshold to 0.1.

## Mean Depth Per Individual

```{r ind-depth}
ind_depth <- read_delim("usub_vcf_subset.idepth", delim = "\t",
                         col_names = c("ind", "nsites", "depth"), skip = 1)

a <- ggplot(ind_depth, aes(depth)) +
  geom_histogram(fill = "dodgerblue1", colour = "black", alpha = 0.3)
a + theme_light()

summary(ind_depth$depth)
```

> Sequencing depth ranges from 12 to 348 — a bit of spread there.

## Proportion of Missing Data Per Individual

```{r ind-missingness}
ind_miss <- read_delim("usub_vcf_subset.imiss", delim = "\t",
                        col_names = c("ind", "ndata", "nfiltered", "nmiss", "fmiss"), skip = 1)

a <- ggplot(ind_miss, aes(fmiss)) +
  geom_histogram(fill = "dodgerblue1", colour = "black", alpha = 0.3)
a + theme_light()

summary(ind_miss$fmiss)
```

> Missingness per individual ranges from 0.03 to 0.20.
